---
title: "Exploratory Analysis"
author: "Rohan Krishnan"
date: "2024-05-04"
output: html_document
---

## Load data & basic info

Look at basic data information (variable types, missing values, unique values, data structure).

```{r}
#Load EDA libraries
library(ggplot2)
library(ggcorrplot)
library(tidyverse)
library(gridExtra)

#Load data
data <- read.csv("~/Documents/GitHub/song-popularity/data/train.csv")

#Dimension of data
dim(data)

#Quick look at data
head(data)

#List variable type for each column -- album_name, track_name, track_genre are characters
for (i in 1:ncol(data)){
  print(paste0(colnames(data[i]),": ", typeof(data[,i])))
}

#Count missing values for each column -- No missing values
print("NA's by column")
for (i in 1:ncol(data)){
  print(paste0(colnames(data[i]), ": ", sum(is.na(data[,i]))))
}

#Count unique values for each column -- explicity is binary, key is categorical, mode is binary, time_signature is categorical, track_genre is categorical 
print("Distinct values by column")
for (i in colnames(data)){
  x = n_distinct(data[i])
  print(paste(i," distinct values: ", x))
}

#We don't need id, album_name, or track_name for EDA
data <- data %>%
  select(-id, -album_name, - track_name)

ncol(data)
```

## Basic EDA

```{r}
#Graph distributions of each variable
for(i in 1:ncol(data)){
  if(typeof(data[,i]) == "character"){
    print("column is character type")
  } else if(typeof(data[,i]) == "logical"){
    print("column is logical type")
  } else{
    p1 <- ggplot(data, aes(x = data[,i])) + 
          geom_density() + 
          labs(x = colnames(data[i]))
    p2 <- ggplot(data, aes(x = data[,i])) + 
          geom_histogram() + 
          labs(x = colnames(data[i]))
    gridExtra::grid.arrange(p1, p2)
  }
}

#Look at outliers with boxplots
for(i in 1:ncol(data)){
  if(typeof(data[,i]) == "character"){
    print(paste(colnames(data[i])," is character type"))
  } else if(typeof(data[,i]) == "logical"){
    print(paste(colnames(data[i])," is logical type"))
  } else{
    p1 <- ggplot(data, aes(x = data[,i])) + 
          geom_boxplot() + 
          labs(x = colnames(data[i]))
    p2 <- ggplot(data, aes(x = data[,i])) + 
          geom_density() + 
          labs(x = colnames(data[i]))
    gridExtra::grid.arrange(p1, p2)
  }
}
```

## Further Exploration of Differences Between Genres

```{r}
#Look at how many songs are represented by each genre
data %>%
  group_by(track_genre) %>%
  summarize(count = n())

#Look at how many songs are explicit vs not
data %>%
  group_by(explicit) %>%
  summarize(count = n()) #There are only 50 (less than 5% of the data) explicit obs

#Look at how many songs have each tempo -- again, some categories with under 5% of obs (0 and 2)
data %>%
  group_by(time_signature) %>%
  summarize(count = n())

#Look at mean values across genres
genre_means <- data %>%
  group_by(track_genre) %>%
  summarise(across(everything(),mean)); 

genre_means %>%
  t() %>%
  as.data.frame()

#Look at boxplots + densities between genres
for(i in 1:ncol(data)){
  if(typeof(data[,i]) == "character"){
    print(paste(colnames(data[i])," is character type"))
  } else if(typeof(data[,i]) == "logical"){
    print(paste(colnames(data[i])," is logical type"))
  } else{
    p1 <- ggplot(data, aes(x = data[,i], color = track_genre)) + 
          geom_boxplot() + 
          labs(x = colnames(data[i]))
    p2 <- ggplot(data, aes(x = data[,i], color = track_genre)) + 
          geom_density() + 
          labs(x = colnames(data[i]))
    gridExtra::grid.arrange(p1, p2)
  }
}

#Look at boxplots + densities between time signatures
for(i in 1:ncol(data)){
  if(typeof(data[,i]) == "character"){
    print(paste(colnames(data[i])," is character type"))
  } else if(typeof(data[,i]) == "logical"){
    print(paste(colnames(data[i])," is logical type"))
  } else{
    p1 <- ggplot(data, aes(x = data[,i], color = as.factor(time_signature))) + 
          geom_boxplot() + 
          labs(x = colnames(data[i]))
    p2 <- ggplot(data, aes(x = data[,i], color = as.factor(time_signature))) + 
          geom_density() + 
          labs(x = colnames(data[i]))
    gridExtra::grid.arrange(p1, p2)
  }
}

#Create correlation matrix - first remove categorical variables

newdata <- data %>%
  select(-time_signature, -explicit, -track_genre)


#newdata$num_genre <- data$track_genre

#unique(newdata$num_genre)

#for(i in nrow(newdata)){
#  if(newdata$num_genre == "pop"){
#  newdata$num_genre == 1
#  }else if(newdata$num_genre == "rock"){
#    newdata$num_genre == 2
#  } else{
#      newdata$num_genre == 3
#    }
#}

#Get correlations and display in matrix
cor <- cor(newdata)
ggcorrplot(cor, title = "Correlation Matrix", tl.cex = 7, legend.title = "Correlation", colors = c("#5386E4", "white", "salmon"))
```

## Main Takeaways

There appears to be significant differences between the popularity of songs between genres. There are also some missing categories in the `time_signature` variable. Before building models, we should preprocess the data and remove any obs with `time_signature` equal to 0 (only 1 obs) or 2 (no obs). We should also convert `explicit`, `time_signature`, `key`, and `track_genre` into factor variables. Finally, we should remove the `id`, `track_name`, and `album_name` variables as they are not variables that we want our models to take into account when trying to predict a song's popularity.
